### 2013-03-14 doc/sosp13/communication.tex

At the upper layer, Grappa implements asynchronous active messages
\cite{vonEicken92}. Each message consists of a function pointer, an
optional argument payload, and an optional data payload. When a task
sends a message, the message is copied to a send queue associated with
the message's destination and the task continues execution.

Grappa's lower networking layer aggregates the upper layer's messages
to improve performance. Commodity networks including infiniband
achieves their peak bisection bandwidth \emph{only} when the packet
sizes are relatively large---on the order of multiple kilobytes. The
reason for this discrepancy is the combination of overheads associated
with handling each packet (in terms of bytes that form the actual
packet, processing time at the card and processing on the
CPU within the driver stack). Our measurements confirm manufacturers
published data~\cite{infinibandbandwidth}, that with this packet size
the bisection bandwidth is only a small fraction, less than
3\% of the peak bisection bandwidth.

In our experiments the vast majority of requests were smaller than
44 bytes, far too small to make
efficient use of the network. To make the best use of the network, we
must convert our small messages into large ones.
When a task sends a message,
it is not immediately sent, but rather placed in a queue specific to the
destination.

There are three situations in which a queue of aggregated messages is
sent. First, each queue has a message size threshold of 4096 bytes, chosen to give reasonable network performance. If the size in
bytes of a queue is above the threshold, the contents of the queue are
sent immediately. Second, each queue has a wait time threshold
($\approx${1ms}). If the oldest message in a queue has been waiting
longer than this threshold, the contents of the queue are sent
immediately, even if the queue size is lower than the message size
threshold.  Third, queues may be explicitly flushed in situations where
the programmer wants to minimize the latency of a message at the cost of
bandwidth utilization.

@@

>>>>

At the upper (user-level) layer, Grappa implements asynchronous active
messages~\cite{vonEicken92}. Each message consists of a function pointer, an
optional argument payload, and an optional data payload. When a task sends a
message, the message is copied to a send queue associated with the message's
destination and the task continues execution.

Grappa’s lower networking layer aggregates the upper layer’s messages to
improve performance. Commodity networks including infiniband achieves their
peak bisection bandwidth only when the packet sizes are relatively large ---
on the order of multiple kilobytes. The reason for this discrepancy is the
combination of overheads associated with handling each packet (in terms of
bytes that form the actual packet, processing time at the card and processing
on the CPU within the driver stack).

In our experiments the vast majority of requests were smaller than 44 bytes.
Our measurements confirm manufacturers' published data [15]; with 44-byte
packets, the available bisection bandwidth is only a small fraction (3\%) of
the peak bisection bandwidth. To make the best use of the network, we must
convert our small messages into large ones. When a task sends a message, it is
not immediately sent, but rather placed in a queue specific to the
destination.

There are three situations in which a queue of aggregated messages is sent:
(1) Each queue has a message size threshold of 4096 bytes, chosen to give
reasonable network performance. If the size in bytes of a queue is above the
threshold, the contents of the queue are sent immediately. (2) Each queue
has a wait time threshold ($\approx${1ms}). If the oldest message in a queue
has been waiting longer than this threshold, the contents of the queue are
sent immediately, even if the queue size is lower than the message size
threshold. (3) Queues may be explicitly flushed in situations where the
programmer wants to minimize the latency of a message at the cost of bandwidth
utilization.

@@

\TODO{We should expand on the zero-copy aggregation stuff?}

### 2013-03-14 doc/sosp13/memory.tex

\begin{figure}[htbp]
  \begin{center}
    \begin{minipage}{\columnwidth}
	\small
	\begin{tabular}{l}
      	\texttt{\scriptsize global\_address global\_malloc( size )} \\
      	\texttt{\scriptsize global\_free( global\_address )} \\ 
      	Allocation in the global heap \\ \\
      	\texttt{\scriptsize delegate\_read( global\_address, local\_var )}  \\
      	\texttt{\scriptsize delegate\_write( global\_address, local\_var )} \\
      	\texttt{\scriptsize delegate\_cas( global\_address, local\_var )} \\
      	\texttt{\scriptsize delegate\_fetch\_inc( global\_address, local\_var )} \\ 
      	Memory operations at the home core of a global address \\ \\
      	\texttt{\scriptsize cache\_acquire( global\_address, local\_buf, \{RO,RW,WO\})} \\
      	\texttt{\scriptsize cache\_release( global\_address, local\_buf )} \\ 
		Perform cache operations to acquire/release global data.  \\
		Acquire copies all data to local node and returns a pointer. \\ 	
		Release optionally copies data back to global memory. \\
	\end{tabular}
      \caption{\label{fig:accessing-memory} Grappa API for memory accesses}     \end{minipage}
  \end{center}
\end{figure}

@@

\textit{Delegate operations.} When the access pattern has low-locality,
it is more efficient to modify the data on its home core rather than
bringing a copy to the requesting core and returning it after
modification. Delegate operations provide this capability. Applications
can dispatch computation to be performed on individual machine-word
sized chunks of global memory to the memory system itself (e.g.,
\emph{fetch-and-add}).  Delegate operations, proposed in
\cite{Nelson:hotpar11} and \cite{delegated:oopsla11}, are also the
primary synchronization method in Grappa.

@@

>>>>

\begin{figure}[htbp]
  \begin{center}
    \begin{minipage}{\columnwidth}
	\small
	\begin{tabular}{l}
      	\texttt{\scriptsize global\_address global\_malloc( size )} \\
      	\texttt{\scriptsize global\_free( global\_address )} \\ 
      	Allocation in the global heap \\ \\
      	\texttt{\scriptsize delegate\_read( global\_address, local\_var )}  \\
      	\texttt{\scriptsize delegate\_write( global\_address, local\_var )} \\
      	\texttt{\scriptsize delegate\_cas( global\_address, local\_var )} \\
      	\texttt{\scriptsize delegate\_fetch\_inc( global\_address, local\_var )} \\ 
      	Memory operations at the home core of a global address \\ \\
      	\texttt{\scriptsize cache\_acquire( global\_address, local\_buf, \{RO,RW,WO\})} \\
      	\texttt{\scriptsize cache\_release( global\_address, local\_buf )} \\ 
		Perform cache operations to acquire/release global data.  \\
		Acquire copies all data to local node and returns a pointer. \\ 	
		Release optionally copies data back to global memory. \\
	\end{tabular}
      \caption{\label{fig:accessing-memory} Grappa API for memory accesses. \TODO{Update table with new calls. And async calls for delegates.}}     \end{minipage}
  \end{center}
\end{figure}

@@

\textit{Delegate operations.} When the access pattern has low-locality,
it is more efficient to modify the data on its home core rather than
bringing a copy to the requesting core and returning it after
modification. Delegate operations provide this capability. Applications
can dispatch computation to be performed on individual machine-word
sized chunks of global memory to the memory system itself (e.g.,
\emph{fetch-and-add}).  Delegate operations, proposed in
\cite{Nelson:hotpar11} and \cite{delegated:oopsla11}, are used to implement synchronization in Grappa.

@@

\TODO{Talk about synchronous and asynchronous delegate operations. Talk about global completion object. Should we have a whole section on synchronization?}

### 2013-03-14 doc/sosp13/relatedwork.tex

Grappa is built on many existing ideas in programming languages, systems and
architecture. In this section we discuss related frameworks and key enabling
technologies that Grappa builds upon.

\paragraph{Comparable frameworks} Distributed graph processing
frameworks like Pregel~\cite{pregel:2010} and Distributed
GraphLab~\cite{distgraphlab:vldb12} share similar goals as
Grappa. Pregel adopts a bulk-synchronous parallel (BSP) execution
model, which makes it inefficient on workloads that could prioritize
vertices. GraphLab, on the other hand, schedules vertex computations
individually, allowing prioritization, which gives faster convergence
in a variety of iterative algorithms.  GraphLab, however, imposes a
rigid computation model where programmers must express computation as
transformations on a vertex and its edge list only, with information only
from adjacent vertexes. Pregel is only slightly less
restrictive, as the input data can be any vertex in the graph.  Grappa
also supports dynamic parallelism with asynchronous execution, but
parallelism is expressed as tasks or loop iterations, which is a far
more general programming model for irregular computation tasks.

\paragraph{Global memory} Grappa includes a custom implementation of a
software distributed shared memory (DSM) system. Many traditional
software DSM systems are page based~\cite{Treadmarks,munin} and aim to
hide the fact that they are built in software from applications by
exploiting the processor's paging mechanisms, therefore relying
heavily on locality. Instead, Grappa, like other partitioned global
address space (PGAS) models, implements its DSM at the language,
rather than system level. Languages such as
Chapel~\cite{Chamberlain:2007}, X10~\cite{X10:2005}, and
UPC~\cite{upc:2005} make accesses to shared structures look like
normal memory references. As we describe later, Grappa chooses a
middle ground, where global addresses are explicit in the API and
local accesses are emitted conventionally by the compiler.  While
Grappa's DSM system is conceptually similar to prior work, its
implementation is tuned for irregular computations.  Past DSM work,
being page-based, could exploit the RDMA capabilities of network
hardware to move large page-sized blocks of data from node to node.
In our experience, when these networks move small blocks of data (a
few bytes), only a fraction of the available bandwidth is achieved.
In addition, the DSM system in Grappa ends up being tightly coupled to
the task scheduler in order to overlap long latency memory operations
with useful computation.  For these two reasons it became necessary to
build a new DSM system specifically for Grappa.

@@

\paragraph{Multithreading} Grappa uses multithreading to tolerate
memory latency. This is a well known technique. Hardware
implementations include the Tera MTA~\cite{tera:mta1}, Cray
XMT~\cite{feo:xmt}, Simultaneous multithreading~\cite{tullsen:smt},
MIT Alewife~\cite{agarwal:alewife}, Cyclops~\cite{almasi:cyclops}, and
even GPUs~\cite{gpus}. As we describe in this paper, Grappa use a
lightweight user-mode task scheduler to multiplex \emph{thousands\/} of
tasks on a single processing core. The large number of tasks is
required because of the extremely high internode latency Grappa is
mitigating.  Grappa's task library employs several optimizations: an
extremely fast task switch, a small task size, and judicious use of
hardware prefetching to bring task state into the cache long before
that task is actually scheduled.

>>>>

% Grappa is built on many existing ideas in programming languages, systems and
% architecture. In this section we discuss related frameworks and key enabling
% technologies that Grappa builds upon.

\paragraph{Distributed graph processing} While \Grappa is a general runtime
system for any large-scale concurrent application, one of its killer
applications is graph analysis. Therefore, distributed graph processing
frameworks like Pregel~\cite{pregel:2010} and Distributed
GraphLab~\cite{distgraphlab:vldb12} are related. Pregel adopts a
bulk-synchronous parallel (BSP) execution model, which makes it inefficient on
workloads that could prioritize vertices. GraphLab, on the other hand,
schedules vertex computations individually, allowing prioritization, which
gives faster convergence in a variety of iterative algorithms. GraphLab,
however, imposes a rigid computation model where programmers must express
computation as transformations on a vertex and its edge list only, with
information only from adjacent vertexes. Pregel is only slightly less
restrictive, as the input data can be any vertex in the graph. Grappa also
supports dynamic parallelism with asynchronous execution, but parallelism is
expressed as tasks or loop iterations, which is a far more general programming
model for irregular computation tasks.

\paragraph{Global memory} Grappa includes support for global memory namespace,
effectively a custom implementation of a software distributed shared memory
(DSM) system.

Many traditional software DSM systems are page based~\cite{Treadmarks,munin}
and aim to hide the fact that they are built in software from applications by
exploiting the processor's paging mechanisms, therefore relying heavily on
locality. Instead, Grappa, like partitioned global address space (PGAS)
models, implements its DSM at the language, rather than the system level.
Languages such as Chapel~\cite{Chamberlain:2007}, X10~\cite{X10:2005}, and
UPC~\cite{upc:2005} make accesses to shared structures look like normal
references. As we describe later, Grappa chooses a middle ground, where global
addresses are explicit in the API and local accesses are emitted
conventionally by the compiler. Also, since past DSM work was mostly
page-based, it could exploit network's RDMA support to move large page-sized
blocks of data from node to node. However, when commodity networks move small
blocks of data (a few bytes), only a fraction of the available bandwidth is
achieved.

In summary, while Grappa's DSM system is conceptually similar to prior work,
it builds latency tolerance from the ground up (message aggregation, constant
overlap of long latency memory operations with computation, etc.), so it can
better support irregular computations with little locality.

% Importantly, global memory support in Grappa
% ends up being tightly coupled to the task scheduler in order to overlap long
% latency memory operations with useful computation. For these two reasons it
% became necessary to build a new DSM system specifically for Grappa.

@@

\paragraph{Multithreading} Grappa uses multithreading to tolerate memory
latency. This is a well known technique. Hardware implementations include the
Tera MTA~\cite{tera:mta1}, Cray XMT~\cite{feo:xmt}, Simultaneous
multithreading~\cite{tullsen:smt}, MIT Alewife~\cite{agarwal:alewife},
Cyclops~\cite{almasi:cyclops}, and even GPUs~\cite{gpus}. As we describe in
this paper, Grappa implements its own software-based multithreading with a
lightweight user-mode task scheduler to multiplex \emph{thousands\/} of tasks
on a single processing core. The large number of tasks is required because of
the to tolerate very high inter-node communication latency. Grappa's task
library employs several optimizations: an extremely fast task switch, a small
task size, and judicious use of hardware prefetching to bring task state into
the cache long before that task is actually scheduled. The main difference between Grappa's support for lightweight threads and prior work such as QThread~\cite{qt} and Capriccio~\cite{capriccion} is the pervasive use of prefetching (with a large number of active tasks, not even their contexts fit in cache!).
